manuscript_outline

### 2025-07-17 docs/manuscript/paper.Rmd
changes for EcoApps

---
title: ""
output:  
  bookdown::pdf_document2:
    toc: false
    number_sections: false
header-includes:
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{setspace}
   - \doublespacing
csl: methods-in-ecology-and-evolution.csl
bibliography: references.bib
---

@@

```{r tail-frog, fig.width=8, fig.height=3, fig.align='center', fig.cap="Custom hatching phenology model for coastal tailed frogs (Ascaphus truei).Panel A represents the model fit and raw data used to generate the effective value model and panel B are the effective values for daily temperatures between 6 and 20 °C. A dashed line with a 0.01 increase for every degree increase in included for reference."}
# Brown 1975, Comparitive Biochemistry and Physicology Part A: Physiology
# https://www.sciencedirect.com/science/article/pii/030096297590033X

@@

p1 + p2 + plot_annotation(tag_levels = 'A')
```

@@

| Class | Order | Genus | Species | Study |
|:---|:---|:---|:---|:---|
| Amphibia | Anura | *Lithobates* | *L. sylvaticus* | @moore1939 |
|  |  |  | *L. pipiens* |  |
|  |  |  | *L. clamitans* |  |
|  |  |  | *L. palustris* |  |
|  |  | *Ascaphus* | *A. truei* | @brown1975 |
|  | Urodela | *Ambystoma* | *A. gracile* | @brown1976 |
| Reptilia | Squamata | *Sceloporus* | *S. undulatus* | @angilletta2000 |
|  |  | *Podarcis* | *P. muralis* | @vandamme1992 |
|  | Testudines | *Mauremys* | *M. reevesii* | @du2007 |
|  |  |  | 181 species | 141 studies in @while2018 |
| Insecta | Plecoptera | *Nemurella* | *N. pictetii* | @brittain1978, @elliott1984 |
|  |  | *Capnia* | *C. atra* | @brittain1984 |
|  |  | *Capnia* | *C. bifrons* | @elliott1986 |
|  |  | *Mesocapnia* | *M. oenone* | @brittain1984 |
|  |  | *Taeniopteryx* | *T. nebulosa* | @brittain1977 |
|  | Coleoptera | *Colaphellus* | *C. bowringi* | @tang2017 |
|  |  |  | 18 species | Developmental equations in @pritchard1987 |
| Malacostraca | Decapoda | *Pontastacus* | *P. leptodactylus* | @aydin2004 |
| Copepoda | Calanoida | 10 genera | 28 species | @forster2011 |
|  | Cyclopoida | *Limnoithona* | *L. tetraspina* |  |
|  |  | *Oithona* | *O. davisae* |  |
|  | Harpacticoida | *Microsetella* | *M. norvegica* |  |
|  |  | *Mesochra* | *M.lilljeborgi* |  |
|  | Poecilostomatoida | *Oncaea* | *O. venusta* |  |
| Cephalopoda | Octopoda | *Octopus* | *O. vulgaris* | @márquez2021 |
|  |  |  | *O. mimus* |  |
|  | Myopsida | *Loligo* | *L. vulgaris* |  |
|  |  |  | *L. reynaudii* |  |
|  | Oegopsida | *Illex* | *I. coindetii* |  |
|  |  |  | *I. illecebrosus* |  |
|  |  | *Todarodes* | *T.pacificus* |  |
|  |  | *Ommastrephes* | *O. bartramii* |  |
| Asteroidea | Valvatida | *Odontaster* | *O. meridionalis* | @hoegh-guldberg1995 |
|  |  |  | *O. validus* |  |
|  |  | *Asterina* | *A. miniata* |  |
|  |  | *Acanthaster* | *A. planci* |  |

@@

```{r frog-preds, fig.height=3 , fig.width=8, fig.align='center', fig.cap="Modeled changes in developmental timing for coastal tailed frogs (Ascaphus truei) from 1991 to 2020. Panel A: Marginal mean trend across all sites showing a significant decline in days to hatching over time. Panel B: Site-specific (COMID) trends in predicted days to hatch, colored by mountain range. Estimates are based on the best fitting model, with stream temperatures from @siegel2023 and developmental timing modeled using hatchR. Shaded ribbons represent 95% confidence intervals."}

@@

p.01 + p.02 + plot_annotation(tag_levels = 'A')

@@

```{r litho-pheno, fig.height=8 , fig.width=8, fig.align='center', fig.cap="Custom hatch timing models developed for four North American *Lithobates* frog species. Models are parameterized from @moore1939 and phenology is predicted using two synthetic temperature regimes: Panel A shows predictions under a cooler regime (mean = 16 °C), while Panel B shows predictions under a warmer regime (mean = 24 °C). Panel C displays the effective value curves for each species—linearized approximations of thermal reaction norms. The crossover in rank order between Panels A and B illustrates genotype-by-environment interactions, reflecting species-specific thermal niches and breeding strategies."}
### all the Rana's below refer to taxonomy from the paper, they are now genus Lithobates

@@

(p2.1.a + p2.1.b) / p2.2 + plot_annotation(tag_levels = "A")
```

@@

```{r beetles, fig.height=3 , fig.width=5, fig.align='center', fig.cap="Effective value models for five populations of cabbage beetles (Colaphellus bowringi, females) along a latitudinal gradient in China [@tang2017]. Models were parameterized from common garden experiments under constant temperatures. The southernmost population (Xiushui County) shows a consistently faster developmental rate than populations from more northerly sites, especially at colder temperatures. This pattern is indicative of cogradient variation, where both genetic and environmental effects reinforce the observed phenotypic trend."}
tang_temps <- c(16, 19, 22, 24, 26, 28)

@@

ef_vals_pops |>
  tibble() |>
  ggplot(aes(x = temperature, y = effective_value, color = beetle_pops)) +
  geom_line() +
  geom_point() +
  # geom_abline(intercept = 0, slope = .01, linetype = "dashed") +
  labs(x = "Daily Average Temperature (°C)", y = "Effective Value") +
  scale_color_brewer(palette = "Dark2", name = "Beetle Populations") +
  theme_classic() +
  theme(legend.position = c(0.25, 0.75))
```

@@

# Bibliography

>>>>

---
title: ""
output:  
  bookdown::word_document2:
    toc: false
    number_sections: false
header-includes:
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{setspace}
   - \doublespacing
csl: ecological-applications.csl
bibliography: references.bib
---

# A generalizable tool for predicting developmental phenology for wild poikilotherms.

Morgan M. Sparks^1,3^, Brian C. Leavell^2^, Bryan M. Maitland^1^

^1^Rocky Mountain Research Station, U.S. Forest Service, Boise, Idaho, USA

^2^Smithsonian Tropical Research Institute, Apartado, Balboa, Ancón, Panamá

^3^Corresponding author, [Morgan.Sparks\@usda.gov](mailto:Morgan.Sparks@usda.gov){.email}

Journal: Ecology Applications

Type: Method

### Open Research Policy

**hatchR** is an open-source software environment distributed via CRAN and additional details are hosted at software website (<https://bmait101.github.io/hatchR/>). Reproducible methods for this manuscript are available at the manuscript github (<https://github.com/morgan-sparks/hatchR_nonfish>) in the `paper.Rmd` file. The repository will become public and publicly archived with a DOI at acceptance of the manuscript.

@@

```{r tail-frog}
# Brown 1975, Comparitive Biochemistry and Physicology Part A: Physiology
# https://www.sciencedirect.com/science/article/pii/030096297590033X

@@

fig1 <- p1 + p2 + plot_annotation(tag_levels = 'A', title = "Figure 1")

@@

@@

```{r frog-preds, fig.height=3 , fig.width=8, fig.align='center', fig.cap=""}

@@

fig2 <- p.01 + p.02 + plot_annotation(tag_levels = 'A', title = "Figure 2")

ggsave("./figures/Figure_2.pdf", fig2, height = 4, width = 8, units = "in", dpi = 1000)

@@

```{r litho-pheno, fig.height=8 , fig.width=8, fig.align='center', fig.cap=""}
### all the Rana's below refer to taxonomy from the paper, they are now genus Lithobates

@@

fig3 <- (p2.1.a + p2.1.b) / p2.2 + plot_annotation(tag_levels = "A", title = "Figure 3")

ggsave("./figures/Figure_3.pdf", fig3, height = 8, width = 8, units = "in", dpi = 1000)
```

@@

```{r beetles, fig.height=3 , fig.width=5, fig.align='center', fig.cap=""}
tang_temps <- c(16, 19, 22, 24, 26, 28)

@@

fig4 <- ef_vals_pops |>
  tibble() |>
  ggplot(aes(x = temperature, y = effective_value, color = beetle_pops)) +
  geom_line() +
  geom_point() +
  # geom_abline(intercept = 0, slope = .01, linetype = "dashed") +
  labs(x = "Daily Average Temperature (°C)", y = "Effective Value", title = "Figure 4") +
  scale_color_brewer(palette = "Dark2", name = "Beetle Populations") +
  theme_classic() +
  theme(legend.position = c(0.25, 0.75))

ggsave("./figures/Figure_4.pdf", fig4, height = 4, width = 8, units = "in", dpi = 1000)
```

@@

# Acknowledgements

We thank N. Karraker, D. Pilliod, M. Hayes, R.B. Bury, and B. Norman for their help in translating sites for the tailed frog example to locations that could be aligned with COMIDs. D. Horan helped organize GPS locations into COMIDs. The views expressed in this manuscript are those of the authors and do not necessarily represent the views or policies of USFS . Any mention of trade names, products, or services does not imply an endorsement by the U.S. government or USFS. USFS does not endorse any commercial products, services or enterprises.

# Author Contributions

Morgan Sparks and Bryan Maitland developed the software and conceived the manuscript. All authors collected data and wrote and edited the manuscript. This manuscript describes an application for software for poikilotherms. As such, it has no specific locale for data collection, however significant effort was made to represent a globally and taxonomically diverse set of species for case studies and model sources.

# Conflict of Interest Statement

The authors declare no known conflicts of interest.

# References

::: {#refs}
:::

# Tables

| Class | Order | Genus | Species | Study |
|:--------------|:--------------|:--------------|:--------------|:--------------|
| Amphibia | Anura | *Lithobates* | *L. sylvaticus* | @moore1939 |
|  |  |  | *L. pipiens* |  |
|  |  |  | *L. clamitans* |  |
|  |  |  | *L. palustris* |  |
|  |  | *Ascaphus* | *A. truei* | @brown1975 |
|  | Urodela | *Ambystoma* | *A. gracile* | @brown1976 |
| Reptilia | Squamata | *Sceloporus* | *S. undulatus* | @angilletta2000 |
|  |  | *Podarcis* | *P. muralis* | @vandamme1992 |
|  | Testudines | *Mauremys* | *M. reevesii* | @du2007 |
|  |  |  | 181 species | 141 studies in @while2018 |
| Insecta | Plecoptera | *Nemurella* | *N. pictetii* | @brittain1978, @elliott1984 |
|  |  | *Capnia* | *C. atra* | @brittain1984 |
|  |  | *Capnia* | *C. bifrons* | @elliott1986 |
|  |  | *Mesocapnia* | *M. oenone* | @brittain1984 |
|  |  | *Taeniopteryx* | *T. nebulosa* | @brittain1977 |
|  | Coleoptera | *Colaphellus* | *C. bowringi* | @tang2017 |
|  |  |  | 18 species | Developmental equations in @pritchard1987 |
| Malacostraca | Decapoda | *Pontastacus* | *P. leptodactylus* | @aydin2004 |
| Copepoda | Calanoida | 10 genera | 28 species | @forster2011 |
|  | Cyclopoida | *Limnoithona* | *L. tetraspina* |  |
|  |  | *Oithona* | *O. davisae* |  |
|  | Harpacticoida | *Microsetella* | *M. norvegica* |  |
|  |  | *Mesochra* | *M.lilljeborgi* |  |
|  | Poecilostomatoida | *Oncaea* | *O. venusta* |  |
| Cephalopoda | Octopoda | *Octopus* | *O. vulgaris* | @márquez2021 |
|  |  |  | *O. mimus* |  |
|  | Myopsida | *Loligo* | *L. vulgaris* |  |
|  |  |  | *L. reynaudii* |  |
|  | Oegopsida | *Illex* | *I. coindetii* |  |
|  |  |  | *I. illecebrosus* |  |
|  |  | *Todarodes* | *T.pacificus* |  |
|  |  | *Ommastrephes* | *O. bartramii* |  |
| Asteroidea | Valvatida | *Odontaster* | *O. meridionalis* | @hoegh-guldberg1995 |
|  |  |  | *O. validus* |  |
|  |  | *Asterina* | *A. miniata* |  |
|  |  | *Acanthaster* | *A. planci* |  |

# Figure Captions

Figure 1: Custom hatching phenology model for coastal tailed frogs (Ascaphus truei).Panel A represents the model fit and raw data used to generate the effective value model and panel B are the effective values for daily temperatures between 6 and 20 °C. A dashed line with a 0.01 increase for every degree increase in included for reference.

Figure 2: Modeled changes in developmental timing for coastal tailed frogs (Ascaphus truei) from 1991 to 2020. Panel A: Marginal mean trend across all sites showing a significant decline in days to hatching over time. Panel B: Site-specific (COMID) trends in predicted days to hatch, colored by mountain range. Estimates are based on the best fitting model, with stream temperatures from @siegel2023 and developmental timing modeled using hatchR. Shaded ribbons represent 95% confidence intervals.

Figure 3: Custom hatch timing models developed for four North American *Lithobates* frog species. Models are parameterized from @moore1939 and phenology is predicted using two synthetic temperature regimes: Panel A shows predictions under a cooler regime (mean = 16 °C), while Panel B shows predictions under a warmer regime (mean = 24 °C). Panel C displays the effective value curves for each species—linearized approximations of thermal reaction norms. The crossover in rank order between Panels A and B illustrates genotype-by-environment interactions, reflecting species-specific thermal niches and breeding strategies.

Figure 4: Effective value models for five populations of cabbage beetles (Colaphellus bowringi, females) along a latitudinal gradient in China [@tang2017]. Models were parameterized from common garden experiments under constant temperatures. The southernmost population (Xiushui County) shows a consistently faster developmental rate than populations from more northerly sites, especially at colder temperatures. This pattern is indicative of cogradient variation, where both genetic and environmental effects reinforce the observed phenotypic trend.

# Figures

```{r  fig.width=8, fig.height=4, fig.align='center', fig.cap=""}
fig1
```

```{r fig.width=8, fig.height=4, fig.align='center', fig.cap=""}
fig2
```

```{r fig.width=8, fig.height=8, fig.align='center', fig.cap=""}
fig3
```

```{r fig.width=8, fig.height=4, fig.align='center', fig.cap=""}
fig4
```
